Recalibrating Gravitational Wave Phenomenological Waveform Model

### 2024-01-05 src/tex/ms.tex
change color of comments

\definecolor{rb4}{HTML}{27408B}
\newcommand{\kw}[1]{{\color{rb4}[KW: #1 ]}}
\definecolor{cyan}{HTML}{0097A7}
\newcommand{\kl}[1]{{\color{cyan}[KL: #1 ]}}
\definecolor{rr}{RGB}{173, 37, 37}
\newcommand{\te}[1]{{\color{rr}[TE: #1 ]}}
\definecolor{cyan}{HTML}{0097A7}
\newcommand{\rr}[1]{{\color{cyan}[#1]}}

@@

Many data analysis tasks in gravitational wave (GW) astrophysics, such
as \rr{matched} filtering~\citep{owen1996search, owen1999matched} and parameter 
estimation~\citep{Dax:2021tsq, Christensen:2022bxb, dynesty, Islam:2022afg, 
Romero-Shaw:2020owr, zackay2018relative, Ashton:2018jfp}, rely upon accurate waveform
models. Since the generation of numerical relativity (NR) waveforms is
prohibitively expensive, the community has constructed waveform approximants
that can be evaluated much faster. There are three families of commonly used GW
approximants: the effective-one-body (EOB) \citep{ossokine2020multipolar,
cotesta2020frequency, taracchini2014effective}, NR
surrogate \citep{islam2022surrogate, varma2019surrogate, varma2019surrogate2},
and phenomenological (Phenom) models \citep{husa2016frequency,
khan2016frequency, garcia2020multimode, pratten2021computationally}. While the
detailed construction of each model is different, they all have a set of
internal parameters that can be calibrated to NR waveforms. The quality of the
waveform model is therefore determined by the ansatz used and the accuracy of the
calibrated parameters.

@@

Aligned-spin, frequency-domain waveform models (such as IMRPhenomD) can be written as a
combination of amplitude and phase functions ($A$ and $\phi$ respectively):
\begin{align}\label{eq:}
	h(f,\theta,\Lambda) = A(f,\theta,\Lambda)e^{-i\phi(f,\theta,\Lambda^i)}\,,
\end{align}
where $f$ is the frequency, $\theta$ are the intrinsic parameters of the binary, and $\Lambda^i$ \rr{($i=1,2,\cdots,19$) is a set of 19 additional
parameters} which will be discussed below. 
The phase and amplitude functions are then split into three sections which represent the
inspiral, intermediate, and merger-ringdown (MR) parts of the waveform. 
During inspiral, $A$ and $\phi$ are known analytically from post-Newtonian (PN) theory;
IMRPhenomD uses the TaylorF2 model \citep{Buonanno:2009zt, Arun:2004hn} which is known up 
to 3.5PN order \kl{augmented with the next four higher order PN terms}.
To model the intermediate and MR regions, IMRPhenomD (and all waveforms in the IMRPhenom family)
uses a series of parameterizations\footnote{
	The parameterizations for both the amplitude and phase functions can be found in \citep{khan2016frequency}.
} 
which depend purely on $\Lambda^i$ and can be calibrated to numerical relativity (NR) simulations.
The three sections are then \textit{stitched} together using step functions.
Importantly, the parameterizations are chosen such that they can be made $\mathcal{C}^1$ continuous at the
boundary between each section.

\begin{figure}
	\script{loss.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/loss.pdf}
	\caption{\rr{Average Loss $\mathcal{L}_{\mathrm{ave}}$} against the number of iterations. The vertical
	dotted line indicates the minimum of the validation loss, which is where we stop
	the optimization.}
	\label{fig:loss}
\end{figure}

@@

In order to optimize the coefficients, we need to define a loss function that
quantifies the difference between the waveform model and the target NR simulations which we want to match.
Here we adopt a quantity commonly used in GW physics called the \textit{mismatch} function \citep{owen1996search, husa2016frequency}. 
It is defined as
\begin{align} \label{eq:mismatch}
	\mathcal{M}(h_1, h_2)=1-\max_{t_0, \phi_0}\langle \hat{h}_1, \hat{h}_2\rangle,
\end{align}
where $h_{1,2}$ are the two GW waveforms we are comparing, and $t_0$ and $\phi_0$
are a relative time and phase shift respectively. 
The inner product, $\langle h_1, h_2 \rangle$, is defined as 
\begin{align}\label{eq:inner_prod}
	\langle h_1, h_2 \rangle = 4\Re\int_{f_{\mathrm{min}}}^{f_{\mathrm{max}}}\frac{h_1(f)h_2^{\ast}(f)}{S_n(f)}\,df,
\end{align}
where $\hat{h}=h/\sqrt{\langle h, h \rangle}$ is the normalized GW strain,
$S_n(f)$ is the power spectral density (PSD), and $f_{\mathrm{max}}$ ($f_{\mathrm{min}}$) is
the maximum (minimum) frequency for the integration.
We note here that the mismatch can be viewed as \rr{a} mean square error (MSE) between the
two waveforms.

@@

Since we wish to optimize the model over the whole parameter space,
we need to compare multiple model generated waveforms with NR waveforms.
However, the mismatch is only defined for two input waveforms at a particular set of intrinsic parameters.
We therefore define the loss function as an average of training waveforms in two ways,
the simple average of mismatches and the normalized average of mismatches,  
\begin{align}\label{eqn:loss}
	\mathcal{L}_{\mathrm{ave}}&=\frac{1}{N}\sum_{i=1}^N\mathcal{M}_i \\
	\mathcal{L}_{\mathrm{norm}}&=\frac{1}{N}\sum_{i=1}^N\frac{\mathcal{M}_i}{\mathcal{M}_{i,\mathrm{ini}}},
\end{align}	
where $\mathcal{M}_i$ represents the mismatch of an individual training waveform,
$\mathcal{M}_{i,\mathrm{ini}}$ represents the mismatch \rr{between original IMRPhenomD waveforms and NR waveforms}, and $N$ is the total number of training waveforms.

The two loss functions are chosen to give different 
behavior during the optimization process.
The simple average, \rr{$\mathcal{L}_{\mathrm{ave}}$}, serves as the simplest choice of loss function,
but is prone to be dominated by a single point in parameter space with a large mismatch.
Other points with smaller mismatches would be insignificant comparatively, and might not be able 
to improve under such a loss function.
Alternatively, the normalized average, $\mathcal{L}_{\mathrm{norm}}$, eliminates the aforementioned issue
by encouraging the waveform to improve at each training point at a similar rate.
The ratio in $\mathcal{L}_{\mathrm{norm}}$ will therefore remain approximately the same for each training point.
Conversely, \rr{$\mathcal{L}_{\mathrm{ave}}$} allows the loss function to automatically adjust
and preferentially optimize the largest mismatches, encouraging the waveform to have similar mismatches everywhere.
In this paper, we show the results of using both loss functions 
and examine the differences between them.

>>>>

\definecolor{rb4}{HTML}{27408B}
\newcommand{\kw}[1]{{\color{rb4}[KW: #1 ]}}
\definecolor{cyan}{HTML}{0097A7}
\newcommand{\kl}[1]{{\color{cyan}[KL: #1 ]}}
\definecolor{rr}{RGB}{173, 37, 37}
\newcommand{\te}[1]{{\color{rr}[TE: #1 ]}}
\definecolor{cyan}{HTML}{0097A7}
\newcommand{\corr}[1]{{\color{red}[corr: #1 ]}}

@@

Many data analysis tasks in gravitational wave (GW) astrophysics, such
as \corr{matched} filtering~\citep{owen1996search, owen1999matched} and parameter 
estimation~\citep{Dax:2021tsq, Christensen:2022bxb, dynesty, Islam:2022afg, 
Romero-Shaw:2020owr, zackay2018relative, Ashton:2018jfp}, rely upon accurate waveform
models. Since the generation of numerical relativity (NR) waveforms is
prohibitively expensive, the community has constructed waveform approximants
that can be evaluated much faster. There are three families of commonly used GW
approximants: the effective-one-body (EOB) \citep{ossokine2020multipolar,
cotesta2020frequency, taracchini2014effective}, NR
surrogate \citep{islam2022surrogate, varma2019surrogate, varma2019surrogate2},
and phenomenological (Phenom) models \citep{husa2016frequency,
khan2016frequency, garcia2020multimode, pratten2021computationally}. While the
detailed construction of each model is different, they all have a set of
internal parameters that can be calibrated to NR waveforms. The quality of the
waveform model is therefore determined by the ansatz used and the accuracy of the
calibrated parameters.

@@

Aligned-spin, frequency-domain waveform models (such as IMRPhenomD) can be written as a
combination of amplitude and phase functions ($A$ and $\phi$ respectively):
\begin{align}\label{eq:}
	h(f,\theta,\Lambda) = A(f,\theta,\Lambda)e^{-i\phi(f,\theta,\Lambda^i)}\,,
\end{align}
where $f$ is the frequency, $\theta$ are the intrinsic parameters of the binary, and $\Lambda^i$ \corr{($i=1,2,\cdots,19$) is a set of 19 additional
parameters} which will be discussed below. 
The phase and amplitude functions are then split into three sections which represent the
inspiral, intermediate, and merger-ringdown (MR) parts of the waveform. 
During inspiral, $A$ and $\phi$ are known analytically from post-Newtonian (PN) theory;
IMRPhenomD uses the TaylorF2 model \citep{Buonanno:2009zt, Arun:2004hn} which is known up 
to 3.5PN order \kl{augmented with the next four higher order PN terms}.
To model the intermediate and MR regions, IMRPhenomD (and all waveforms in the IMRPhenom family)
uses a series of parameterizations\footnote{
	The parameterizations for both the amplitude and phase functions can be found in \citep{khan2016frequency}.
} 
which depend purely on $\Lambda^i$ and can be calibrated to numerical relativity (NR) simulations.
The three sections are then \textit{stitched} together using step functions.
Importantly, the parameterizations are chosen such that they can be made $\mathcal{C}^1$ continuous at the
boundary between each section.

\begin{figure}
	\script{loss.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/loss.pdf}
	\caption{\corr{Average Loss $\mathcal{L}_{\mathrm{ave}}$} against the number of iterations. The vertical
	dotted line indicates the minimum of the validation loss, which is where we stop
	the optimization.}
	\label{fig:loss}
\end{figure}

@@

In order to optimize the coefficients, we need to define a loss function that
quantifies the difference between the waveform model and the target NR simulations which we want to match.
Here we adopt a quantity commonly used in GW physics called the \textit{mismatch} function \citep{owen1996search, husa2016frequency}. 
It is defined as
\begin{align} \label{eq:mismatch}
	\mathcal{M}(h_1, h_2)=1-\max_{t_0, \phi_0}\langle \hat{h}_1, \hat{h}_2\rangle,
\end{align}
where $h_{1,2}$ are the two GW waveforms we are comparing, and $t_0$ and $\phi_0$
are a relative time and phase shift respectively. 
The inner product, $\langle h_1, h_2 \rangle$, is defined as 
\begin{align}\label{eq:inner_prod}
	\langle h_1, h_2 \rangle = 4\Re\int_{f_{\mathrm{min}}}^{f_{\mathrm{max}}}\frac{h_1(f)h_2^{\ast}(f)}{S_n(f)}\,df,
\end{align}
where $\hat{h}=h/\sqrt{\langle h, h \rangle}$ is the normalized GW strain,
$S_n(f)$ is the power spectral density (PSD), and $f_{\mathrm{max}}$ ($f_{\mathrm{min}}$) is
the maximum (minimum) frequency for the integration.
We note here that the mismatch can be viewed as \corr{a} mean square error (MSE) between the
two waveforms.

@@

Since we wish to optimize the model over the whole parameter space,
we need to compare multiple model generated waveforms with NR waveforms.
However, the mismatch is only defined for two input waveforms at a particular set of intrinsic parameters.
We therefore define the loss function as an average of training waveforms in two ways,
the simple average of mismatches and the normalized average of mismatches,  
\begin{align}\label{eqn:loss}
	\mathcal{L}_{\mathrm{ave}}&=\frac{1}{N}\sum_{i=1}^N\mathcal{M}_i \\
	\mathcal{L}_{\mathrm{norm}}&=\frac{1}{N}\sum_{i=1}^N\frac{\mathcal{M}_i}{\mathcal{M}_{i,\mathrm{ini}}},
\end{align}	
where $\mathcal{M}_i$ represents the mismatch of an individual training waveform,
$\mathcal{M}_{i,\mathrm{ini}}$ represents the mismatch \corr{between original IMRPhenomD waveforms and NR waveforms}, and $N$ is the total number of training waveforms.

The two loss functions are chosen to give different 
behavior during the optimization process.
The simple average, \corr{$\mathcal{L}_{\mathrm{ave}}$}, serves as the simplest choice of loss function,
but is prone to be dominated by a single point in parameter space with a large mismatch.
Other points with smaller mismatches would be insignificant comparatively, and might not be able 
to improve under such a loss function.
Alternatively, the normalized average, $\mathcal{L}_{\mathrm{norm}}$, eliminates the aforementioned issue
by encouraging the waveform to improve at each training point at a similar rate.
The ratio in $\mathcal{L}_{\mathrm{norm}}$ will therefore remain approximately the same for each training point.
Conversely, \corr{$\mathcal{L}_{\mathrm{ave}}$} allows the loss function to automatically adjust
and preferentially optimize the largest mismatches, encouraging the waveform to have similar mismatches everywhere.
In this paper, we show the results of using both loss functions 
and examine the differences between them.
